---
title: "Metabolomics Analysis: *S. mutans* UA159, lrgAB, and atlA Mutants Under Aerobic/Anaerobic Growth"
subtitle: "Ahn et al. study — amino acids, acylcarnitines, organic acids, NAD/adenylate/CoA pools"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
    fig_width: 8
    fig_height: 5.5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                       fig.align = "center")
```

# 1. Overview

This report parses the two source workbooks describing the study and its
results:

* **`Ahn_study.xls`** — Metabolomics Workbench-style study design (`Study Design` sheet):
  which strain, sample type, growth medium, and growth condition each numbered
  sample corresponds to.
* **`Ahn_Results.xlsx`** — one "Results" sheet per metabolite class (amino
  acids, acylcarnitines, organic acids, NAD-related metabolites, adenylates
  ATP/ADP/AMP, and acyl-CoAs), each giving metabolite abundance across the
  numbered samples, plus a matching "Additional Info" sheet with annotation
  (retention index, PubChem/KEGG IDs).

The study compares three *Streptococcus mutans* strains — **UA159 (wild
type)**, **lrgAB mutant**, and **atlA mutant** — grown **aerobically** or
**anaerobically**, profiling both **whole cells** and, for UA159/lrgAB only,
the **culture supernatant ("Media")**. Three biological replicates were
collected per condition.

Two design quirks in the source data drive how the analysis below is
organized, and both are read directly from which sample columns exist in
each `Results` sheet rather than assumed:

1. **Not every strain has every panel.** Amino acids, organic acids, NAD
   metabolites, and adenylates (ATP/ADP/AMP) were only run on UA159 and
   lrgAB — the atlA mutant simply has no columns in those sheets. Acylcarnitines
   are the *only* panel that includes all three strains. This is why plots
   below sometimes legitimately show two strains and sometimes three — Section
   5.1 makes the exact coverage explicit so it's never a silent omission.
2. **Whole cells and supernatant are different matrices, not the same pool
   in two forms** — cell pellets are reported in `nmol / mg protein`, media
   in `uM`, over different (and only partly overlapping) strain/sample sets. They
   are never pooled into one "value" for statistics or PCA; every comparison
   below keeps `Matrix` (Cells vs. Media) as an explicit grouping variable,
   and Sections 6.3 and 8 place the two side by side, panel-for-panel, per
   metabolite so intracellular pool and secretion can be read together
   without conflating their scales.

Values flagged `BLOQ` (below limit of quantitation), `N.D.` (not detected),
or `N/A` (unsuitable sample) are treated as censored/missing rather than
zero, which is standard practice for targeted metabolomics panels.

```{r libraries}
library(readxl)   # read .xls / .xlsx
library(dplyr)
library(tidyr)
library(stringr)
library(tibble)
library(purrr)
library(forcats)
library(ggplot2)
library(scales)

theme_set(theme_bw(base_size = 12))
```

# 2. File paths

Edit these two lines to point at your local copies of the workbooks (they
can be `.xls`/`.xlsx` interchangeably — `readxl` handles both).

```{r paths}
study_path   <- "Ahn_study.xls"
results_path <- "Ahn_Results.xlsx"
```

# 3. Import sample metadata (Study Design)

```{r sample-metadata}
clean_id <- function(x) {
  x <- trimws(as.character(x))
  sub("\\.0$", "", x)   # readxl sometimes round-trips whole numbers as "7.0"
}

sd_raw <- read_excel(study_path, sheet = "Study Design", skip = 2, col_names = FALSE)

sample_meta <- sd_raw %>%
  transmute(
    Sample_ID        = clean_id(...3),
    Strain           = as.character(...4),
    Sample_Type      = as.character(...5),
    Growth_Media     = as.character(...6),
    Growth_Condition = as.character(...7)
  ) %>%
  filter(!is.na(Sample_ID), Sample_ID != "") %>%
  mutate(
    # media-only blanks (C1-1..C1-3) carry "N/A" instead of a real strain/condition
    Strain           = ifelse(Sample_Type == "Media only", "Media blank", Strain),
    Growth_Condition = ifelse(Sample_Type == "Media only", "n/a", Growth_Condition),
    Replicate = {
      num_id <- suppressWarnings(as.numeric(Sample_ID))
      case_when(
        str_detect(Sample_ID, "^C1-") ~ str_extract(Sample_ID, "[0-9]+$"),
        num_id <= 10 ~ "1",
        num_id <= 20 ~ "2",
        TRUE          ~ "3"
      )
    },
    Strain = factor(Strain, levels = c("UA159 wild type", "lrgAB mutant",
                                        "atlA mutant", "Media blank")),
    Growth_Condition = factor(Growth_Condition,
                               levels = c("anaerobically", "aerobically", "n/a"))
  )

knitr::kable(sample_meta, caption = "Sample -> strain / condition mapping")
```

# 4. Generic reader for the "Results" sheets

Every `Results *` sheet has the same shape once you find the header row (the
row starting with "Metabolite name"): an id block (`Metabolite name`,
optionally `InChI Key`, `Units of Measurement`) followed by one column per
sample. This single function tidies any of them into long format and flags
censored values.

```{r results-reader}
read_results_sheet <- function(path, sheet) {
  raw <- read_excel(path, sheet = sheet, col_names = FALSE, .name_repair = "minimal")

  hdr_row <- which(apply(raw, 1, function(r)
    any(str_detect(tolower(trimws(as.character(r))), "^metabolite name$"))))[1]
  stopifnot(!is.na(hdr_row))

  header <- as.character(unlist(raw[hdr_row, ]))
  header[is.na(header)] <- paste0("V", seq_along(header))[is.na(header)]

  dat <- raw[(hdr_row + 1):nrow(raw), ]
  colnames(dat) <- header
  dat <- dat %>% filter(!is.na(`Metabolite name`))

  id_cols     <- intersect(c("Metabolite name", "InChI Key", "Units of Measurement"),
                            colnames(dat))
  sample_cols <- setdiff(colnames(dat), id_cols)

  # coerce all sample columns to character first: some columns are pure
  # numeric and some are numeric+"BLOQ" text, which otherwise clash when
  # pivoted together
  dat <- dat %>% mutate(across(all_of(sample_cols), as.character))

  dat %>%
    pivot_longer(all_of(sample_cols), names_to = "Sample_ID", values_to = "value_raw") %>%
    mutate(
      Sample_ID = clean_id(Sample_ID),
      value_num = suppressWarnings(as.numeric(value_raw)),
      Flag = case_when(
        value_raw %in% c("BLOQ")                          ~ "BLOQ",
        str_detect(coalesce(value_raw, ""), "N\\.D")        ~ "ND",
        str_detect(coalesce(value_raw, ""), "N/A")          ~ "N/A",
        is.na(value_num) & !is.na(value_raw)                ~ "other",
        TRUE                                                 ~ "OK"
      ),
      Units = if ("Units of Measurement" %in% colnames(dat)) `Units of Measurement` else NA_character_
    ) %>%
    rename(Metabolite = `Metabolite name`) %>%
    select(Metabolite, Sample_ID, Units, value_num, Flag)
}
```

# 5. Import every metabolite panel and join metadata

```{r import-all}
sheet_map <- tribble(
  ~sheet,                ~Class,                   ~Matrix,
  "Results AA- Cells",   "Amino acids",            "Cells",
  "Results AA- Media",   "Amino acids",            "Media",
  "Results AC- Cells",   "Acylcarnitines",         "Cells",
  "Results OA- Cells",   "Organic acids",          "Cells",
  "Results OA- Media",   "Organic acids",          "Media",
  "Results NAD",         "NAD metabolites",        "Cells",
  "Results ATP",         "Energy (ATP/ADP/AMP)",   "Cells",
  "Results CoA",         "Acyl-CoA",               "Cells"
)

results_long <- pmap_dfr(sheet_map, function(sheet, Class, Matrix) {
  read_results_sheet(results_path, sheet) %>% mutate(Class = Class, Matrix = Matrix)
})

results_annot <- results_long %>% left_join(sample_meta, by = "Sample_ID")

cat("Total metabolite x sample measurements:", nrow(results_annot), "\n")
cat("Rows that failed to match a sample:", sum(is.na(results_annot$Strain)), "\n")
```

## 5.1 Data coverage by strain and compartment

This is the direct answer to "why do some plots only show two strains":
count how many samples of each strain actually exist in each panel /
compartment. A `0` means that combination genuinely was not run — not that
the code dropped it.

```{r coverage-table}
coverage <- results_annot %>%
  distinct(Class, Matrix, Strain, Sample_ID) %>%
  count(Class, Matrix, Strain, name = "n_samples") %>%
  tidyr::pivot_wider(names_from = Strain, values_from = n_samples, values_fill = 0) %>%
  arrange(Class, Matrix)

knitr::kable(coverage, caption = "Number of samples per Class x Matrix x Strain")
```

**Reading this table:** the atlA mutant (`atlA mutant` column) is non-zero
*only* for Acylcarnitines/Cells — every other panel is UA159 + lrgAB only,
which is a property of the original study, not a filtering choice made
here. Likewise `Media` (supernatant) rows exist only for Amino acids and
Organic acids — the other panels (acylcarnitines, NAD metabolites,
adenylates, acyl-CoAs) were only ever measured intracellularly, so there is
no supernatant version of those plots to make.

## 5.2 Data-quality summary

```{r qc-summary, fig.height=4}
qc <- results_annot %>%
  count(Class, Matrix, Flag) %>%
  group_by(Class, Matrix) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ungroup()

ggplot(qc, aes(x = Class, y = pct, fill = Flag)) +
  geom_col() +
  facet_wrap(~Matrix, scales = "free_x") +
  coord_flip() +
  scale_fill_manual(values = c(OK = "#2c7fb8", BLOQ = "#f2b134", ND = "#d95f02",
                                "N/A" = "grey60", other = "black")) +
  labs(title = "Share of censored vs. quantified values per metabolite panel",
       x = NULL, y = "% of measurements")
```

Many acylcarnitine species are below the limit of quantitation across the
board — the plots below automatically keep only the species that were
actually detected somewhere, per compartment.

# 6. Amino acids: intracellular pool and secretion

Amino acids were profiled in both compartments (UA159 and lrgAB only — see
5.1), so every comparison here is done twice: once for the intracellular
pool (`Cells`, nmol/mg protein) and once for what ends up in the supernatant
(`Media`, uM). The two are kept as separate panels throughout because their
units and sample sets differ; they are never summed or averaged together.

```{r total-aa-pool, fig.height=4.5, fig.width=9}
total_aa <- results_annot %>%
  filter(Class == "Amino acids", !is.na(value_num)) %>%
  group_by(Sample_ID, Matrix, Strain, Growth_Condition, Replicate) %>%
  summarise(total = sum(value_num), .groups = "drop") %>%
  mutate(Matrix = factor(Matrix, levels = c("Cells", "Media")))

ggplot(total_aa, aes(Growth_Condition, total, fill = Strain)) +
  geom_boxplot(position = position_dodge(0.8), alpha = 0.75, outlier.shape = NA) +
  geom_point(position = position_jitterdodge(jitter.width = 0.08, dodge.width = 0.8),
             size = 2) +
  facet_wrap(~Matrix, scales = "free_y",
             labeller = labeller(Matrix = c(Cells = "Intracellular pool (nmol/mg protein)",
                                             Media = "Secreted into media (uM)"))) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Total amino acid pool: intracellular vs. secreted",
       y = NULL, x = NULL)
```

## 6.1 Individual amino acids, intracellular pool

```{r aa-panel, fig.width=11, fig.height=10}
aa_cells <- results_annot %>%
  filter(Class == "Amino acids", Matrix == "Cells") %>%
  group_by(Metabolite) %>%
  filter(any(Flag == "OK")) %>%
  ungroup()

ggplot(aa_cells, aes(Growth_Condition, value_num, fill = Strain)) +
  geom_boxplot(position = position_dodge(0.8), alpha = 0.75, outlier.shape = NA) +
  geom_point(position = position_jitterdodge(jitter.width = 0.08, dodge.width = 0.8),
             size = 0.9) +
  facet_wrap(~Metabolite, scales = "free_y", ncol = 4) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Intracellular amino acids by strain and growth condition",
       y = "nmol / mg protein", x = NULL) +
  theme(legend.position = "top", strip.text = element_text(size = 8))
```

## 6.2 Individual amino acids, secreted into media

```{r aa-panel-media, fig.width=11, fig.height=10}
aa_media <- results_annot %>%
  filter(Class == "Amino acids", Matrix == "Media") %>%
  group_by(Metabolite) %>%
  filter(any(Flag == "OK")) %>%
  ungroup()

ggplot(aa_media, aes(Growth_Condition, value_num, fill = Strain)) +
  geom_boxplot(position = position_dodge(0.8), alpha = 0.75, outlier.shape = NA) +
  geom_point(position = position_jitterdodge(jitter.width = 0.08, dodge.width = 0.8),
             size = 0.9) +
  facet_wrap(~Metabolite, scales = "free_y", ncol = 4) +
  scale_fill_manual(values = c("UA159 wild type" = "#66C2A5", "lrgAB mutant" = "#FC8D62",
                                "Media blank" = "#8DA0CB")) +
  labs(title = "Amino acids secreted into culture media by strain and growth condition",
       subtitle = "The 'n/a' column is the uninoculated media blank (baseline)",
       y = "uM", x = NULL) +
  theme(legend.position = "top", strip.text = element_text(size = 8))
```

## 6.3 Intracellular pool vs. secretion, side by side

The most direct way to see whether a strain/condition difference in the
cell pellet shows up as a difference in what's exported: each row below is
one amino acid, with its intracellular panel on the left and its secreted
panel on the right — each on its own y-axis, since the units aren't
comparable. `drop = FALSE` keeps every metabolite's row aligned to two
columns even where one side had no quantifiable data, so a metabolite two
rows apart in Section 6.1 stays lined up with its own media panel here.

```{r aa-compartment-compare, fig.width=9, fig.height=13}
aa_key <- c("Alanine", "Glutamine", "Glutamic Acid", "Serine", "Glycine", "Aspartic Acid")

aa_compare <- results_annot %>%
  filter(Class == "Amino acids", Metabolite %in% aa_key,
         Strain %in% c("UA159 wild type", "lrgAB mutant")) %>%
  mutate(Matrix = factor(Matrix, levels = c("Cells", "Media")),
         Metabolite = factor(Metabolite, levels = aa_key))

ggplot(aa_compare, aes(Growth_Condition, value_num, fill = Strain)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA, alpha = 0.75) +
  geom_point(position = position_jitterdodge(jitter.width = 0.08, dodge.width = 0.8), size = 1) +
  facet_wrap(~ Metabolite + Matrix, scales = "free_y", ncol = 2, drop = FALSE,
             labeller = labeller(Matrix = c(Cells = "Intracellular (nmol/mg protein)",
                                             Media = "Secreted (uM)"),
                                  .multi_line = FALSE)) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Amino acids: intracellular pool (left) vs. secretion into media (right)",
       subtitle = paste("Shown for a representative subset -",
                         "swap `aa_key` above for any of the metabolites in 6.1/6.2"),
       x = NULL, y = NULL)
```

# 7. Acylcarnitines (the only panel that includes the atlA mutant)

```{r acylcarnitines, fig.width=9, fig.height=6}
ac_detected <- results_annot %>%
  filter(Class == "Acylcarnitines", Matrix == "Cells") %>%
  group_by(Metabolite) %>%
  filter(any(Flag == "OK")) %>%
  ungroup()

ggplot(ac_detected, aes(Growth_Condition, value_num, fill = Strain)) +
  geom_boxplot(position = position_dodge(0.8), alpha = 0.75, outlier.shape = NA) +
  geom_point(position = position_jitterdodge(jitter.width = 0.08, dodge.width = 0.8),
             size = 1.3) +
  facet_wrap(~Metabolite, scales = "free_y") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Acylcarnitine species detected above LOQ (Cells, all 3 strains)",
       y = "nmol / mg protein", x = NULL)
```

# 8. Organic acids (central carbon / TCA-cycle intermediates)

Like amino acids, organic acids were measured both intracellularly and in
the supernatant (UA159 and lrgAB only), and the same metabolite names
recur in both sheets. Here the media panel also includes the **media-blank
control** (`Media blank`, plotted at the `n/a` growth-condition position) —
the baseline level in uninoculated BHI, useful for telling net secretion
apart from what was already in the broth (e.g. citrate, malate, succinate
are all *higher* in the blank than in inoculated media, i.e. partially
consumed by the bacteria rather than secreted).

```{r organic-acids, fig.width=10, fig.height=6}
oa_cells <- results_annot %>%
  filter(Class == "Organic acids", Matrix == "Cells") %>%
  group_by(Metabolite) %>%
  filter(any(Flag == "OK")) %>%
  ungroup()

ggplot(oa_cells, aes(Growth_Condition, value_num, fill = Strain)) +
  geom_boxplot(position = position_dodge(0.8), alpha = 0.75, outlier.shape = NA) +
  geom_point(position = position_jitterdodge(jitter.width = 0.08, dodge.width = 0.8),
             size = 1.3) +
  facet_wrap(~Metabolite, scales = "free_y") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Intracellular organic acids by strain and growth condition",
       y = "nmol / mg protein", x = NULL)
```

## 8.1 Intracellular pool vs. secretion, side by side

```{r oa-compartment-compare, fig.width=9, fig.height=17}
oa_compare <- results_annot %>%
  filter(Class == "Organic acids") %>%
  group_by(Metabolite) %>%
  filter(any(Flag == "OK")) %>%   # keep the metabolite if detected in EITHER compartment
  ungroup() %>%
  mutate(Matrix = factor(Matrix, levels = c("Cells", "Media")),
         Metabolite = factor(Metabolite))

ggplot(oa_compare, aes(Growth_Condition, value_num, fill = Strain)) +
  geom_boxplot(position = position_dodge(0.8), outlier.shape = NA, alpha = 0.75) +
  geom_point(position = position_jitterdodge(jitter.width = 0.08, dodge.width = 0.8), size = 1) +
  facet_wrap(~ Metabolite + Matrix, scales = "free_y", ncol = 2, drop = FALSE,
             labeller = labeller(Matrix = c(Cells = "Intracellular (nmol/mg protein)",
                                             Media = "Secreted (uM)"),
                                  .multi_line = FALSE)) +
  scale_fill_manual(values = c("UA159 wild type" = "#66C2A5", "lrgAB mutant" = "#FC8D62",
                                "Media blank" = "#8DA0CB")) +
  labs(title = "Organic acids: intracellular pool (left) vs. secretion into media (right)",
       subtitle = "'n/a' on the x-axis = uninoculated media blank (Media panels only)",
       x = NULL, y = NULL) +
  theme(legend.position = "top")
```

A blank intracellular panel (e.g. Malate, 3-HBA) means that organic acid
was below the limit of quantitation in whole cells for both strains, even
though it was clearly detectable in the supernatant — it isn't missing
data, the cell pool for those species is just too small/fast-turned-over
to quantify with this assay.

# 9. Adenylate energy charge (ATP, ADP, AMP)

Energy charge is calculated as
$\text{EC} = \dfrac{[\text{ATP}] + 0.5\,[\text{ADP}]}{[\text{ATP}]+[\text{ADP}]+[\text{AMP}]}$.

```{r energy-charge, fig.height=4.5}
adenylates <- results_annot %>%
  filter(Class == "Energy (ATP/ADP/AMP)") %>%
  select(Sample_ID, Metabolite, value_num, Strain, Growth_Condition, Replicate) %>%
  pivot_wider(names_from = Metabolite, values_from = value_num) %>%
  mutate(Energy_Charge = (ATP + 0.5 * ADP) / (ATP + ADP + AMP))

ggplot(adenylates, aes(Growth_Condition, Energy_Charge, fill = Strain)) +
  geom_boxplot(position = position_dodge(0.8), alpha = 0.75, outlier.shape = NA) +
  geom_point(position = position_jitterdodge(jitter.width = 0.08, dodge.width = 0.8),
             size = 2) +
  scale_fill_brewer(palette = "Set2") +
  ylim(0, 1) +
  labs(title = "Adenylate energy charge", y = "Energy charge", x = NULL)
```

# 10. NAD(H)/NADP(H) pool and acyl-CoAs

```{r nad-coa, fig.width=10, fig.height=6}
nad_coa <- results_annot %>%
  filter(Class %in% c("NAD metabolites", "Acyl-CoA")) %>%
  group_by(Metabolite) %>%
  filter(any(Flag == "OK")) %>%
  ungroup()

ggplot(nad_coa, aes(Growth_Condition, value_num, fill = Strain)) +
  geom_boxplot(position = position_dodge(0.8), alpha = 0.75, outlier.shape = NA) +
  geom_point(position = position_jitterdodge(jitter.width = 0.08, dodge.width = 0.8),
             size = 1.3) +
  facet_wrap(~Metabolite, scales = "free_y") +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "NAD-related metabolites and acyl-CoAs (species detected above LOQ)",
       y = "nmol / mg protein", x = NULL)
```

# 11. Multivariate view: intracellular metabolome PCA

This PCA deliberately combines only same-compartment data: all Cells-matrix
panels (amino acids, acylcarnitines, organic acids, NAD metabolites,
adenylates, acyl-CoAs), restricted to UA159 and lrgAB — the two strains with
a complete panel set (Section 5.1) — so every sample has the same features
on the same `nmol/mg protein` scale. `BLOQ`/non-detected values are imputed
at half the minimum observed value for that metabolite (a standard,
conservative approach for left-censored metabolomics data), and any
metabolite never detected in this strain pair is dropped. The parallel
secretome PCA in 11.2 is kept as a *separate* ordination rather than
merged in, because Media is a different matrix on a different unit scale
(uM) with a partly different sample set (it also includes the media-blank
controls) — pooling the two compartments into one PCA would let the unit
scale dominate the ordination rather than the biology.

```{r pca, fig.height=5.5}
cells_full_panel <- results_annot %>%
  filter(Matrix == "Cells", Strain %in% c("UA159 wild type", "lrgAB mutant"))

cells_imputed <- cells_full_panel %>%
  group_by(Class, Metabolite) %>%
  mutate(
    half_min  = suppressWarnings(min(value_num, na.rm = TRUE)) / 2,
    half_min  = ifelse(is.finite(half_min), half_min, NA_real_),
    value_imp = ifelse(is.na(value_num), half_min, value_num)
  ) %>%
  ungroup() %>%
  filter(!is.na(value_imp))

wide <- cells_imputed %>%
  mutate(feat = paste(Class, Metabolite, sep = " | ")) %>%
  distinct(Sample_ID, feat, .keep_all = TRUE) %>%
  select(Sample_ID, feat, value_imp, Strain, Growth_Condition, Replicate) %>%
  pivot_wider(names_from = feat, values_from = value_imp)

meta_cols <- c("Sample_ID", "Strain", "Growth_Condition", "Replicate")
mat <- wide %>% select(-all_of(meta_cols)) %>% as.matrix()
rownames(mat) <- wide$Sample_ID
mat <- mat[, colSums(is.na(mat)) == 0, drop = FALSE]   # keep fully-observed features

pca <- prcomp(mat, center = TRUE, scale. = TRUE)
varexp <- summary(pca)$importance[2, 1:2] * 100

scores <- as_tibble(pca$x[, 1:2], rownames = "Sample_ID") %>%
  left_join(wide %>% select(Sample_ID, Strain, Growth_Condition), by = "Sample_ID")

ggplot(scores, aes(PC1, PC2, color = Strain, shape = Growth_Condition)) +
  geom_point(size = 3.5) +
  scale_color_brewer(palette = "Set1") +
  labs(x = paste0("PC1 (", round(varexp[1], 1), "%)"),
       y = paste0("PC2 (", round(varexp[2], 1), "%)"),
       title = paste0("PCA of intracellular metabolome (n = ", nrow(mat),
                       " samples, ", ncol(mat), " metabolites)"))
```

## 11.1 Heatmap of z-scored metabolite abundance

```{r heatmap, fig.width=10, fig.height=11}
z <- scale(mat)

hc_feat <- hclust(dist(t(z)))
hc_samp <- hclust(dist(z))

z_long <- as_tibble(z, rownames = "Sample_ID") %>%
  pivot_longer(-Sample_ID, names_to = "feat", values_to = "z") %>%
  mutate(Sample_ID = factor(Sample_ID, levels = rownames(z)[hc_samp$order]),
         feat      = factor(feat,      levels = colnames(z)[hc_feat$order]))

ggplot(z_long, aes(Sample_ID, feat, fill = z)) +
  geom_tile() +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b",
                        midpoint = 0, name = "z-score") +
  labs(title = "Hierarchically clustered, z-scored intracellular metabolite levels",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 8) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        axis.text.y = element_text(size = 6),
        panel.grid = element_blank())
```

## 11.2 Secretome (media) PCA

The extracellular counterpart to Section 11: amino acid + organic acid
panels measured in the supernatant, for UA159, lrgAB, and the media-blank
control — imputed and scaled the same way, but never combined with the
Cells-matrix data above.

```{r media-pca, fig.height=5.5}
media_imputed <- results_annot %>%
  filter(Matrix == "Media") %>%
  group_by(Class, Metabolite) %>%
  mutate(
    half_min  = suppressWarnings(min(value_num, na.rm = TRUE)) / 2,
    half_min  = ifelse(is.finite(half_min), half_min, NA_real_),
    value_imp = ifelse(is.na(value_num), half_min, value_num)
  ) %>%
  ungroup() %>%
  filter(!is.na(value_imp))

wide_media <- media_imputed %>%
  mutate(feat = paste(Class, Metabolite, sep = " | ")) %>%
  distinct(Sample_ID, feat, .keep_all = TRUE) %>%
  select(Sample_ID, feat, value_imp, Strain, Growth_Condition) %>%
  pivot_wider(names_from = feat, values_from = value_imp)

mat_media <- wide_media %>% select(-Sample_ID, -Strain, -Growth_Condition) %>% as.matrix()
rownames(mat_media) <- wide_media$Sample_ID
mat_media <- mat_media[, colSums(is.na(mat_media)) == 0, drop = FALSE]

pca_media <- prcomp(mat_media, center = TRUE, scale. = TRUE)
varexp_media <- summary(pca_media)$importance[2, 1:2] * 100

scores_media <- as_tibble(pca_media$x[, 1:2], rownames = "Sample_ID") %>%
  left_join(wide_media %>% select(Sample_ID, Strain, Growth_Condition), by = "Sample_ID")

ggplot(scores_media, aes(PC1, PC2, color = Strain, shape = Growth_Condition)) +
  geom_point(size = 3.5) +
  scale_color_manual(values = c("UA159 wild type" = "#E41A1C", "lrgAB mutant" = "#377EB8",
                                 "Media blank" = "#4DAF4A")) +
  labs(x = paste0("PC1 (", round(varexp_media[1], 1), "%)"),
       y = paste0("PC2 (", round(varexp_media[2], 1), "%)"),
       title = paste0("Secretome PCA (n = ", nrow(mat_media),
                       " samples, ", ncol(mat_media), " metabolites)"),
       subtitle = "Includes the media-blank control as a reference point")
```

The media-blank samples typically separate clearly from inoculated media on
PC1 — that's the net effect of bacterial consumption/secretion on the
supernatant composition, and is a useful sanity check that the assay is
picking up real biology.

# 12. Session info

```{r session-info}
sessionInfo()
```
